---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First unzip the activity.zip file, if activity.csv is not yet in the working directory.
```{r echo=TRUE}
if(!'activity.csv' %in% dir()){
    unzip('activity.zip')
    unlink('activity.zip')}
```
Load the data and show first six rows.
```{r echo=TRUE}
data <- read.csv('activity.csv', header=TRUE, stringsAsFactors=FALSE)
head(data)
```
Transform the date variable into datetime objects with the lubridate package.
```{r echo=TRUE}
library(lubridate)
data$date <- ymd(data$date)
```

## What is mean total number of steps taken per day?
Create a table with the plyr package that sums the number of steps per day. 
```{r echo=TRUE, message=FALSE}
library(plyr)
stepsperday <- ddply(data, .(date), summarize, steps=sum(steps, na.rm=TRUE))
```

Then make a ggplot histogram of the total number of steps taken each day.
```{r echo=TRUE}
library(ggplot2)
ggplot(stepsperday, aes(date, steps)) + geom_histogram(stat="identity") + ggtitle(("Number of steps per day"))
```

Calculate the mean and median total of steps taken per day.
```{r echo=TRUE}
mean(stepsperday$steps)
median(stepsperday$steps)
```

## What is the average daily activity pattern?
Calculate the average number of steps taken in each five number interval (i.e. the same daily interval averaged across all days) and create a time series plot.
```{r echo=TRUE}
averageinterval <- ddply(data, .(interval), summarize, steps=mean(steps, na.rm=TRUE))
ggplot(averageinterval, aes(interval, steps)) + geom_line() + ggtitle("Average steps per 5-minute interval")
```

The 5-minute interval which contains the maximum number of steps on average, across all the days in the dataset:
```{r echo=TRUE}
head(averageinterval[with(averageinterval, order(steps, decreasing=TRUE)),], n=1)  
```

## Imputing missing values
Count the number of missing values in the data set:
```{r echo=TRUE}
sum(is.na(data))          # in total
sum(is.na(data$steps))    # in steps variable
```
Fill in the missing values with the mean value for that interval and add this to a new dataset:
```{r echo=TRUE}
# Subset the missing data
missing <- is.na(data$steps)
missing_entries <- data[missing, ]
# Calculate the means for each interval
meaninterval <- ddply(data, .(interval), summarize, steps=mean(steps, na.rm=TRUE))
# Create a vector with the average step values for all the missing data
imputed_values <- vector()
for(interval in missing_entries$interval){
    new_step <- meaninterval[meaninterval$interval==interval, 'steps']
    imputed_values <- c(imputed_values, new_step)
    }
# Add this vector to a new copy of the dataset
new_data <- data
new_data[missing, 'steps'] <- imputed_values
# Verify that there are no longer any missing data in the new dataset
sum(is.na(new_data))
```
Now that the missing data has been imputed, let's see how the new data differs from the original data. First, by creating a histogram of the total number of steps taken per day:
```{r echo=TRUE}
new_stepsperday <- ddply(new_data, .(date), summarize, steps=sum(steps))
ggplot(new_stepsperday, aes(date, steps)) + geom_histogram(stat="identity") + ggtitle(("Number of steps per day (missing values imputed)"))  
```

Unsurprisingly, the main difference with the histogram above is that there are now fewer days without total steps. Calculating the mean and median total of steps taken per day, we see that both values have increased substantially compared to the previous calculations:  
```{r echo=TRUE}
mean(new_stepsperday$steps)
median(new_stepsperday$steps)
```
## Are there differences in activity patterns between weekdays and weekends?
