---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First unzip the activity.zip file, if activity.csv is not yet in the working directory.
```{r echo=TRUE}
if(!'activity.csv' %in% dir()){
    unzip('activity.zip')
    unlink('activity.zip')}
```
Load the data and show first six rows.
```{r echo=TRUE}
data <- read.csv('activity.csv', header=TRUE, stringsAsFactors=FALSE)
head(data)
```
Transform the date variable into datetime objects with the lubridate package.
```{r echo=TRUE}
library(lubridate)
data$date <- ymd(data$date)
```

## What is mean total number of steps taken per day?
Create a table with the plyr package that sums the number of steps per day. 
```{r echo=TRUE, message=FALSE}
library(plyr)
stepsperday <- ddply(data, .(date), summarize, steps=sum(steps, na.rm=TRUE))
```

Then make a ggplot histogram of the total number of steps taken each day.
```{r echo=TRUE}
library(ggplot2)
ggplot(stepsperday, aes(date, steps)) + geom_histogram(stat="identity") + ggtitle(("Number of steps per day"))
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
